 Total Team Spending and Profitability Results in a Higher League Ranking in the English Premier League

### 2024-04-13 paper/paper.qmd
Commiting all work done to paper.qmd as of 04-13-24.

library(tidyverse)
library(dplyr)
library(corrplot)
library(dplyr)
library(knitr)
library(readr)
library(here)
library(kableExtra)
library(scales)

@@

The resulting dataset for analysis is shown in @tab-6.

@@

# Assigning new column names to the dataframe
colnames(analysis_data) <- c("Team", "Home Matchday Attendance", "Market Value", "Points", "Total Payroll", "Total Transfer Fees")

# Generate the table with adjusted font size to fit the page
kable(analysis_data, booktabs = TRUE, align = c("l", rep("r", 6)), "latex", longtable = TRUE) |>
  kable_styling(latex_options = c("striped", "HOLD_position"), font_size = 8)
```

>>>>

library(tidyverse)
library(dplyr)
library(corrplot)
library(dplyr)
library(knitr)
library(readr)
library(here)
library(kableExtra)
library(scales)
library(ggplot2)

@@

The resulting dataset for analysis is shown in @tbl-6.

@@

# Creating a temporary copy of the dataframe for table presentation
table_data <- analysis_data

# Specifying new column names for the table aesthetics
names(table_data) <- c("Team", "Home Matchday Attendance", "Market Value", "Points", "Total Payroll", "Total Transfer Fees")

# Generate the table with adjusted column names and font size
kable(table_data, booktabs = TRUE, align = c("l", rep("r", 6)), "latex", longtable = TRUE) |>
  kable_styling(latex_options = c("striped", "HOLD_position"), font_size = 8)
```

From here I began examining the relationships between English Premier League teams' league performance (expressed in points accumulated) and various financial indicators during the 2023-2024 season. The following four graphs (@fig-pts_vs_average_home_matchday_attendance, @fig-pts_vs_total_wage_bill, @fig-pts_vs_market_value, and @fig-pts_vs_transfer_fees) presents a scatter plot paired with a linear regression trend line to evaluate the correlation between league points and one of the four key economic metrics presented in @tbl-6.

```{r}
#| warning: false
#| echo: false
#| label: fig-pts_vs_average_home_matchday_attendance
#| fig-cap: Scatter plot illustrating the relationship between total league points and average matchday attendance during the 2023-2024 season. Each point represents a team, plotted with their total points on the x-axis and their average home attendance on the y-axis. The trend line, added using linear regression analysis, suggests a correlation between team performance and matchday turnout. The correlation coefficient of 0.694060419947882 indicates a positive association with moderate correlation, where teams with higher league points tend to have greater average attendance.

# Calculate correlation coefficient for league position and average attendance
cor_attendance <- cor(analysis_data$pts, analysis_data$average_home_matchday_attendance, use = "complete.obs")

# Create a scatter plot for league position and average attendance
ggplot(analysis_data, aes(x = pts, y = average_home_matchday_attendance)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE) +
  theme_minimal() +
  labs(title = "Figure 1: Average Matchday Attendance in Relation 
       to League Position for the 2023-2024 Season",
       subtitle = paste("Correlation coefficient:", cor_attendance),
       x = "Total League Points",
       y = "Average Attendance")
```

```{r}
#| warning: false
#| echo: false
#| label: fig-pts_vs_total_wage_bill
#| fig-cap: Scatter plot depicting the association between total league points and the total payroll for teams in the 2023-2024 season. Each point on the plot corresponds to a team, positioned according to their accumulated points for the season along the x-axis and their respective payroll on the y-axis. The linear regression trend line suggests a correlation between a team’s financial expenditure on payroll and their success in the league. A correlation coefficient of 0.770829037567151 reflects a positive relationship, implying that teams with higher payrolls may be more likely to accumulate greater league points.

# Calculate correlation coefficient for league position and average total wages
cor_wages <- cor(analysis_data$pts, analysis_data$total_wage_bill, use = "complete.obs")

# Create a scatter plot for league position and average total wages
ggplot(analysis_data, aes(x = pts, y = total_wage_bill)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE) +
  theme_minimal() +
  labs(title = "Figure 2: Total Payroll in Relation to League 
       Points for the 2023-2024 Season",
       subtitle = paste("Correlation coefficient:", cor_wages),
       x = "Total League Points",
       y = "Total Payroll")
```

```{r}
#| warning: false
#| echo: false
#| label: fig-pts_vs_market_value
#| fig-cap: Scatter plot demonstrating the correlation between the current market value of teams and their accumulated league points for the 2023-2024 season. The scatter plot places each team by the total points they have secured on the x-axis against their market value on the y-axis. A pronounced upward trend represented by the linear regression line, with a correlation coefficient of 0.849962012423038, indicates a strong positive relationship. This suggests that teams with higher market values tend to amass more league points, underlining a potential link between financial strength and on-field success.

# Calculate correlation coefficient for league position and market value
cor_market <- cor(analysis_data$pts, analysis_data$market_value, use = "complete.obs")

# Create a scatter plot for league position and market value
ggplot(analysis_data, aes(x = pts, y = market_value)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE) +
  theme_minimal() +
  labs(title = "Figure 4: Current Market Value in Relation 
       to League Points for the 2023-2024 Season",
       subtitle = paste("Correlation coefficient:", cor_market),
       x = "Total League Points",
       y = "Total Market Value")
```

```{r}
#| warning: false
#| echo: false
#| label: fig-pts_vs_transfer_fees
#| fig-cap: Scatter plot illustrating the relationship between teams total transfer spending and their corresponding league points for the 2023-2024 season. The scatter plot shows each team as a distinct point, positioned by their total league points on the x-axis and their transfer spending on the y-axis. The trend line, calculated with linear regression, indicates a moderate positive correlation, with a correlation coefficient of approximately 0.417985956450557. This suggests that higher league points tend to coincide with increased transfer spending, although the relationship is not strongly linear.

# Calculate correlation coefficient for league position and total transfer fees
cor_transfers <- cor(analysis_data$pts, analysis_data$transfer_fees, use = "complete.obs")

# Create a scatter plot for league position and average total wages
ggplot(analysis_data, aes(x = pts, y = transfer_fees)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE) +
  theme_minimal() +
  labs(title = "Figure 3: Average Transfer Spending in Relation to 
       League Points for the 2023-2024 Season",
       subtitle = paste("Correlation coefficient:", cor_transfers),
       x = "Total League Points",
       y = "Total Transfer Fees")
```
